---
title: "Imperial Covid19 Code - base.r"
author: "Hamilton Institute, Maynooth"
date: "01/04/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(data.table)
library(lubridate)
library(gdata)
library(EnvStats)
```


This is our attempt to annotate and understand the Covid-19 modelling code made public at <https://github.com/ImperialCollegeLondon/covid19model>. 
Write up is available at <https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-Europe-estimates-and-NPI-impact-30-03-2020.pdf>

### To run the code

Download/clone the master from  <https://github.com/ImperialCollegeLondon/covid19model>. Open rstudio project covid19model.Rproj file in rstudio and run/source base.r file

### This is the code used in base.r

### Reading in the data

Countries are defined and relevant data is read in.

```{r data, eval = FALSE}
### We should add *Ireland* here once we have included enough Irish data 
countries <- c(
  "Denmark",
  "Italy",
  "Germany",
  "Spain",
  "United_Kingdom",
  "France",
  "Norway",
  "Belgium",
  "Austria", 
  "Sweden",
  "Switzerland"
)

### Set up here to use "base" stan model
args = commandArgs(trailingOnly=TRUE)
if(length(args) == 0) {
  args = 'base'
} 
StanModel = args[1]

print(sprintf("Running %s",StanModel))

### Reading all data
### This contains reporting_date, country ID, number of deaths, number of cases and population size in 2018
### Ireland is included here.
d=readRDS('data/COVID-19-up-to-date.rds')

### Get CFR
### This spreadsheet gives the probability of dying given infection for different age categories. 
### Presumably calculated from deaths and cases that are reported. 
### We would need to add a row for *Ireland*.
### Note that only the overall CFR is used not the age based one.
cfr.by.country = read.csv("data/weighted_fatality.csv")
### Notation
cfr.by.country$country = as.character(cfr.by.country[,2])
### Notation
cfr.by.country$country[cfr.by.country$country == "United Kingdom"] = "United_Kingdom"

### Need to include *Ireland* here..date schools/unis closed and other dates of measures implemented.
covariates = read.csv('data/interventions.csv', stringsAsFactors = FALSE)
### 8 interventions from schools closing to self isolation for 11 chosen countries
covariates <- covariates[1:11, c(1,2,3,4,5,6, 7, 8)]

## making all covariates that happen after lockdown to have same date as lockdown
covariates$schools_universities[covariates$schools_universities > covariates$lockdown] 
<- covariates$lockdown[covariates$schools_universities > covariates$lockdown]

covariates$travel_restrictions[covariates$travel_restrictions > covariates$lockdown] 
<- covariates$lockdown[covariates$travel_restrictions > covariates$lockdown] 

covariates$public_events[covariates$public_events > covariates$lockdown] 
<- covariates$lockdown[covariates$public_events > covariates$lockdown]

covariates$sport[covariates$sport > covariates$lockdown] 
<- covariates$lockdown[covariates$sport > covariates$lockdown]

covariates$social_distancing_encouraged[covariates$social_distancing_encouraged > covariates$lockdown] 
<- covariates$lockdown[covariates$social_distancing_encouraged > covariates$lockdown]

covariates$self_isolating_if_ill[covariates$self_isolating_if_ill > covariates$lockdown] 
<- covariates$lockdown[covariates$self_isolating_if_ill > covariates$lockdown]
```

__A note on SI:__
SI is serial interval, defined in the paper as "the time between when a person gets infected and when they subsequently infect other people". The paper says this is g ~ Gamma(6.5,0.62). You get the data below using dgamma(1:100,6.5,0.62). 


    
```{r SI}
### So serial.interal$fit is the probability density and $X is time in days (1 to 100)
serial.interval = read.csv("data/serial_interval.csv") 

### Plot Fig 7 in paper supplemental
plot(serial.interval$X[1:20], serial.interval$fit[1:20], type = "l")
  

```



### Creating data for stan model

```{r standata1, eval = FALSE}
## number of parameters
p <- ncol(covariates) - 1
forecast = 0

## N2 is the number of days to compute the forecast for 
DEBUG = FALSE
if(DEBUG == FALSE) {
  N2 = 75 # Increase this for a further forecast
}  else  {
  ### For faster runs:
  # countries = c("Austria","Belgium") #,Spain")
  N2 = 75
}
# countries = c("Italy","United_Kingdom","Spain","Norway","Austria","Switzerland")

dates = list()
reported_cases = list()
deaths_by_country = list()

### create stan data object to be added to later
stan_data = list(M=length(countries),
                 N=NULL,
                 p=p,
                 x1=poly(1:N2,2)[,1],
                 x2=poly(1:N2,2)[,2],
                 y=NULL,
                 covariate1=NULL,
                 covariate2=NULL,
                 covariate3=NULL,
                 covariate4=NULL,
                 covariate5=NULL,
                 covariate6=NULL,
                 covariate7=NULL,
                 deaths=NULL,
                 f=NULL,
                 N0=6,
                 cases=NULL,
                 LENGTHSCALE=7,
                 SI=serial.interval$fit[1:N2],
                 EpidemicStart = NULL) # N0 = 6 to make it consistent with Rayleigh

for(Country in countries) {
  ### Get CFR
  CFR=cfr.by.country$weighted_fatality[cfr.by.country$country == Country]
  
  covariates1 <- covariates[covariates$Country == Country, 2:8]
  
  ### Get first non-zero cases
  d1=d[d$Countries.and.territories==Country,]
  d1$date = as.Date(d1$DateRep,format='%d/%m/%Y')
  d1$t = decimal_date(d1$date) 
  d1=d1[order(d1$t),]
  index = which(d1$Cases>0)[1]
  index1 = which(cumsum(d1$Deaths)>=10)[1] # also 5
  index2 = index1-30
  
  print(sprintf("First non-zero cases is on day %d, and 30 days before 5 days is day %d",index,index2))
  d1=d1[index2:nrow(d1),]
  stan_data$EpidemicStart = c(stan_data$EpidemicStart,index1+1-index2)
  
  
  for (ii in 1:ncol(covariates1)) {
    covariate = names(covariates1)[ii]
    d1[covariate] <- (as.Date(d1$DateRep, format='%d/%m/%Y') >= as.Date(covariates1[1,covariate]))*1  # should this be > or >=?
  }
  
  dates[[Country]] = d1$date
```

__A note on probability of death depending on no. days since infection:__

The goal below is to calculate `f` which in the paper is denoted $\pi_{s,m}$; the "infection-to-death distribution", on a given day $s$, for country, $m$. This is the weight used in the calculation of the expected number of deaths. $\pi_{s,m}$ is derived from $\pi_m$ (labeled as `convolution` in the code) where

$\pi_m \sim CFR_m*(\Gamma(5.1,0.86) + \Gamma(18.8,0.45))$,

where $CFR_m$ is the probability of dying given infection for different age categories (pre-calculated and read in above).

Then $\pi_m$ is discretized to give $\pi_{s,m}$ which is used in the calculation for the expected number of deaths in the stan model as follows: 

$d_{t,m} = \sum_{\tau = 0}^{t-1}c_{\tau,m}\times\pi_{t-\tau,m}$

The number of deaths at time t  is the sum of the past infections ($c_{\tau,m}$) weighted by their probability of death, where the probability of death depends on the number of days since infection.

__A note on assumptions in the model:__
Do not assume homogeneous attacks across age-groups - they do here per Cathal Walsh.
Instead they do a funky reweighting (middle of page 17) 
They start modelling from 10 deaths as before that probably due to cases from abroad.

```{r standata2, eval = FALSE}
  ### hazard estimation
  N = length(d1$Cases)
  print(sprintf("%s has %d days of data",Country,N))
  forecast = N2 - N
  if(forecast < 0) {
    print(sprintf("%s: %d", Country, N))
    print("ERROR!!!! increasing N2")
    N2 = N
    forecast = N2 - N
  }
  
  h = rep(0,forecast+N) # discrete hazard rate from time t = 1, ..., 100
  if(DEBUG) { # OLD -- but faster for testing this part of the code
    mean = 18.8
    cv = 0.45
    
    for(i in 1:length(h))
      h[i] = (CFR*pgammaAlt(i,mean = mean,cv=cv) - CFR*pgammaAlt(i-1,mean = mean,cv=cv)) / (1-CFR*pgammaAlt(i-1,mean = mean,cv=cv))
  } else { # NEW 
    ### justified in the paper, NC 010420
    mean1 = 5.1; cv1 = 0.86; # infection to onset
    mean2 = 18.8; cv2 = 0.45 # onset to death
    ## assume that CFR is probability of dying given infection
    x1 = rgammaAlt(5e6,mean1,cv1) # infection-to-onset ----> do all people who are infected get to onset? Incubation period, NC 010420
    x2 = rgammaAlt(5e6,mean2,cv2) # onset-to-death
    
    ### Compute the empirical cumulative distribution function     
    ### This is the discretisation of the two gammas in pi_m
    f = ecdf(x1+x2) 
    ### convolution = CFR*(Gamma(5.1,0.86) + Gamma(18.8,0.45))
    convolution = function(u) (CFR * f(u))
    ### This is pi_m in the paper which is defined as "the infection to death distribution" for country m
    ### This is gives a continuous RV but is discretised later which is what f is below 
    ### (confusing given f is also an object above)
    ### This is different to convolution defined in the stan model, NC 010420

    ### Wikipedia definition of convolution! "In mathematics convolution is a mathematical operation on two functions that produces a  third function expressing how the shape of one is modified by the other. The term convolution refers to both the result function and to the process of computing it.""
    
    ### Discretization of pi_m (convolution)
    h[1] = (convolution(1.5) - convolution(0)) 
    for(i in 2:length(h)) {
      h[i] = (convolution(i+.5) - convolution(i-.5)) / (1-convolution(i-.5))
    }

  }
  s = rep(0,N2)
  s[1] = 1 
  for(i in 2:N2) {
    s[i] = s[i-1]*(1-h[i-1])
  }
  f = s * h ## f is the weight used in estimating the expected no. of deaths. 
```

```{r}
### Plot Gamma distribution for infection to onset of symptoms and onset to death
mean1 = 5.1; cv1 = 0.86; # infection to onset
plot(dgammaAlt(0:40,mean1,cv1) , type = "l",main="Infection to onset of symptoms")
mean2 = 18.8; cv2 = 0.45 # onset to death
plot(dgammaAlt(0:80,mean2,cv2) , type = "l",main="Onset of symptoms to death")
```
      
      
This is just appending all the data to the `stan data` object. 

```{r standata3, eval = FALSE}
  y=c(as.vector(as.numeric(d1$Cases)),rep(-1,forecast))
  reported_cases[[Country]] = as.vector(as.numeric(d1$Cases))
  deaths=c(as.vector(as.numeric(d1$Deaths)),rep(-1,forecast))
  cases=c(as.vector(as.numeric(d1$Cases)),rep(-1,forecast))
  deaths_by_country[[Country]] = as.vector(as.numeric(d1$Deaths))
  covariates2 <- as.data.frame(d1[, colnames(covariates1)])
  # x=1:(N+forecast)
  covariates2[N:(N+forecast),] <- covariates2[N,]
  
  ## append data
  stan_data$N = c(stan_data$N,N)
  stan_data$y = c(stan_data$y,y[1]) # just the index case!
  # stan_data$x = cbind(stan_data$x,x)
  stan_data$covariate1 = cbind(stan_data$covariate1,covariates2[,1])
  stan_data$covariate2 = cbind(stan_data$covariate2,covariates2[,2])
  stan_data$covariate3 = cbind(stan_data$covariate3,covariates2[,3])
  stan_data$covariate4 = cbind(stan_data$covariate4,covariates2[,4])
  stan_data$covariate5 = cbind(stan_data$covariate5,covariates2[,5])
  stan_data$covariate6 = cbind(stan_data$covariate6,covariates2[,6])
  stan_data$covariate7 = cbind(stan_data$covariate7,covariates2[,7]) 
  stan_data$f = cbind(stan_data$f,f)
  stan_data$deaths = cbind(stan_data$deaths,deaths)
  stan_data$cases = cbind(stan_data$cases,cases)
  
  stan_data$N2=N2
  stan_data$x=1:N2
  if(length(stan_data$N) == 1) {
    stan_data$N = as.array(stan_data$N)
  }
}

stan_data$covariate2 = 0 * stan_data$covariate2 # remove travel bans
stan_data$covariate4 = 0 * stan_data$covariate5 # remove sport

#stan_data$covariate1 = stan_data$covariate1 # school closure
stan_data$covariate2 = stan_data$covariate7 # self-isolating if ill
#stan_data$covariate3 = stan_data$covariate3 # public events
# create the `any intervention` covariate
stan_data$covariate4 = 1*as.data.frame((stan_data$covariate1+
                                          stan_data$covariate3+
                                          stan_data$covariate5+
                                          stan_data$covariate6+
                                          stan_data$covariate7) >= 1)
stan_data$covariate5 = stan_data$covariate5 # lockdown
stan_data$covariate6 = stan_data$covariate6 # social distancing encouraged
stan_data$covariate7 = 0 # models should only take 6 covariates

if(DEBUG) {
  for(i in 1:length(countries)) {
    write.csv(
      data.frame(date=dates[[i]],
                 `school closure`=stan_data$covariate1[1:stan_data$N[i],i],
                 `self isolating if ill`=stan_data$covariate2[1:stan_data$N[i],i],
                 `public events`=stan_data$covariate3[1:stan_data$N[i],i],
                 `government makes any intervention`=stan_data$covariate4[1:stan_data$N[i],i],
                 `lockdown`=stan_data$covariate5[1:stan_data$N[i],i],
                 `social distancing encouraged`=stan_data$covariate6[1:stan_data$N[i],i]),
      file=sprintf("results/%s-check-dates.csv",countries[i]),row.names=F)
  }
}

stan_data$y = t(stan_data$y)

```

### Run the model 

```{r runstan, eval = FALSE}
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
m = stan_model(paste0('stan-models/',StanModel,'.stan'))

if(DEBUG) {
  fit = sampling(m,data=stan_data,iter=40,warmup=20,chains=2)
} else { 
  # fit = sampling(m,data=stan_data,iter=4000,warmup=2000,chains=8,thin=4,control = list(adapt_delta = 0.90, max_treedepth = 10))
  fit = sampling(m,data=stan_data,iter=200,warmup=100,chains=4,thin=4,control = list(adapt_delta = 0.90, max_treedepth = 10))
}  
```

### Get Results

```{r res, eval = FALSE}
out = rstan::extract(fit)
prediction = out$prediction
estimated.deaths = out$E_deaths
estimated.deaths.cf = out$E_deaths0

JOBID = Sys.getenv("PBS_JOBID")
if(JOBID == "")
  JOBID = as.character(abs(round(rnorm(1) * 1000000)))
print(sprintf("Jobid = %s",JOBID))

save.image(paste0('results/',StanModel,'-',JOBID,'.Rdata'))

save(fit,prediction,dates,reported_cases,deaths_by_country,countries,estimated.deaths,estimated.deaths.cf,out,covariates,file=paste0('results/',StanModel,'-',JOBID,'-stanfit.Rdata'))

# to visualize results
library(bayesplot)
filename <- paste0('base-',JOBID)
plot_labels <- c("School Closure",
                 "Self Isolation",
                 "Public Events",
                 "First Intervention",
                 "Lockdown", 'Social distancing')
alpha = (as.matrix(out$alpha))
colnames(alpha) = plot_labels
g = (mcmc_intervals(alpha, prob = .9))
ggsave(sprintf("results/%s-covars-alpha-log.pdf",filename),g,width=4,height=6)
g = (mcmc_intervals(alpha, prob = .9,transformations = function(x) exp(-x)))
ggsave(sprintf("results/%s-covars-alpha.pdf",filename),g,width=4,height=6)
mu = (as.matrix(out$mu))
colnames(mu) = countries
g = (mcmc_intervals(mu,prob = .9))
ggsave(sprintf("results/%s-covars-mu.pdf",filename),g,width=4,height=6)
dimensions <- dim(out$Rt)
Rt = (as.matrix(out$Rt[,dimensions[2],]))
colnames(Rt) = countries
g = (mcmc_intervals(Rt,prob = .9))
ggsave(sprintf("results/%s-covars-final-rt.pdf",filename),g,width=4,height=6)
system(paste0("Rscript plot-3-panel.r ", filename,'.Rdata'))
system(paste0("Rscript plot-forecast.r ",filename,'.Rdata')) ## to run this code you will need to adjust manual values of forecast required
```
